Revealing transient strain in geodetic data with Gaussian process regression

### 2017-09-30 RevisedSubmission/hines_hetland_2017_gji.tex
did more

GNSS data records transient displacements as well as other physical and non-physical processes that we are not interested in. We first consider component $i$ of a single GNSS displacement observation made at station $j$, which is located at $\vec{x}^{(j)}$, and time $t$. We describe this observation as a realization of the random variable 
\begin{align}\label{eq:Data}
\begin{split}
d^{(j)}_i(t) = &u_i(\vec{x}^{(j)},t) + \eta_i(\vec{x}^{(j)},t) + a_i^{(j)} + b_i^{(j)}t + c_i^{(j)}\sin(2 \pi t) + \\ 
               &e_i^{(j)}\cos(2 \pi t) + f_i^{(j)}\sin(4 \pi t)  + g_i^{(j)}\cos(4 \pi t), 
\end{split}
\end{align}
where $a_i^{(j)}$ is an offset that is unique for each station, $b_i^{(j)}$ is the secular velocity at $\vec{x}^{(j)}$, and the sinusoids describe seasonal deformation (using units of years for $t$). We use $\eta_i$ to represent all the stochastic noise processes. For example, $\eta_i$ can consist of uncorrelated white noise, temporally correlated noise describing benchmark wobble \citep[e.g.,][]{Wyatt1982,Wyatt1989}, and/or spatially correlated noise describing common mode error \citep[e.g.,][]{Wdowinski1997}. For now, we only make the assumption that $\eta_i$ is a Gaussian process distributed as $\eta_i \sim \mathcal{N}(0,C_{\eta_i})$, and $C_{\eta_i}$ is a spatially and temporally dependent covariance function. We consider the coefficients in eq. (\ref{eq:Data}) to be uncorrelated random variables distributed as $\mathcal{N}(0,\kappa^2)$ in the limit as $\kappa \to \infty$ (i.e., the coefficients have diffuse priors). Of course, the secular velocities, $b_i^{(j)}$, are spatially correlated and we could invoke a tectonic model to form a prior on $b_i^{(j)}$. However, in our application to the Pacific Northwest, we will be using displacement time series which are long enough to sufficiently constrain $b_i^{(j)}$ for each station, avoiding the need to incorporate a prior. Likewise, seasonal deformation is spatially correlated \citep{Dong2002,Langbein2008}, and it may be worth exploring and exploiting such a correlation in a future study.

We move on to consider the column vector of $n$ GNSS displacement observations, $\mitbf{d}^*_i$, where the subscript indicates that we are still only considering component $i$ of displacements. The times and positions for each observation are described by the set $\mitbf{P}$. $\mitbf{d}^*_i$ can be considered a realization of the random vector $\mitbf{d}_i$, which is formed by evaluating eq. (\ref{eq:Data}) at each point in $\mitbf{P}$. To write out $\mitbf{d}_i$ explicitly, we collect the coefficients from eq. (\ref{eq:Data}) into the vector $\mitbf{m}_i$, which contains 6 elements for each of the $m$ GNSS station. We use $\mitbf{G}$ to represent the $n \times 6m$ matrix of corresponding basis functions evaluated at each point in $\mitbf{P}$. We can write $\mitbf{d}_i$ as 
\begin{equation}
\mitbf{d}_i = u_i(\mitbf{P}) + \eta_i(\mitbf{P}) + \mitbf{G}\mitbf{m}_i,
\end{equation}
which is distributed as $\mathcal{N}(\mitbf{0},\mitbf{\Sigma} + \kappa^2\mitbf{G}\mitbf{G}^T)$, where
\begin{equation}\label{eq:Cd}
\mitbf{\Sigma} = C_{u_i}(\mitbf{P},\mitbf{P}) + C_{\eta_i}(\mitbf{P},\mitbf{P}).  
\end{equation}
It should be understood that notation such as $u_i(\mitbf{P})$ and $C_{u_i}(\mitbf{P},\mitbf{P})$ represents the column vector $[u_i(p)]_{p \in \mitbf{P}}$ and the matrix $[C_{u_i}(p,p')]_{(p,p') \in \mitbf{P} \times \mitbf{P}}$, respectively.

We combine the prior on transient displacements with GNSS displacement data to form a posterior estimate of transient displacements. We ignore any covariances between the easting and northing components, which reduces the computational complexity of our analysis because we can determine each component of the posterior independently. However, we are inherently assuming that the principle components of transient displacements and the noise processes are aligned with the cardinal directions, which is admittedly an arbitrary assumption. We then form each component of the posterior, denoted as $\hat{u}_i$, by updating $u_i$ with the fact that $\mitbf{d}^*_i$ has been sampled from the random vector $\mitbf{d}_i$. That is to say $\hat{u}_i = u_i |\mitbf{d}^*_i$. If $\kappa$ is kept finite then, from \citet[sec. 2.2]{Rasmussen2006}, we find that $\hat{u}_i$ is distributed as $\mathcal{N}(\mu_{\hat{u}_i},C_{\hat{u}_i})$, where

@@

\subsection{Outlier detection}\label{sec:Outlier}
In our formulation for estimating transient strain rates, we have assumed that noise in the data vector is normally distributed. This is not an appropriate assumption for GNSS data, which often have more outliers than would be predicted for normally distributed noise. It follows that proposed methods for analyzing GNSS data should be robust against outliers \citep[e.g.,][]{Blewitt2016}. In order to make our estimates of transient strain more robust, we automatically identify and remove outliers in the GNSS data as a pre-processing step.

>>>>

GNSS data records transient displacements as well as other physical and non-physical processes that we are not interested in. We first consider component $i$ of a single GNSS displacement observation made at station $j$, which is located at $\vec{x}^{(j)}$, and time $t$. We describe this observation as a realization of the random variable 
\begin{align}\label{eq:Data}
\begin{split}
d^{(j)}_i(t) = &u_i(\vec{x}^{(j)},t) + \eta_i(\vec{x}^{(j)},t) + a_i^{(j)} + b_i^{(j)}t + c_i^{(j)}\sin(2 \pi t) + \\ 
               &e_i^{(j)}\cos(2 \pi t) + f_i^{(j)}\sin(4 \pi t)  + g_i^{(j)}\cos(4 \pi t), 
\end{split}
\end{align}
where $a_i^{(j)}$ is an offset that is unique for each station, $b_i^{(j)}$ is the secular velocity at $\vec{x}^{(j)}$, and the sinusoids describe seasonal deformation (using units of years for $t$). We use $\eta_i$ to represent all the non-parametric noise processes. For example, $\eta_i$ can consist of uncorrelated white noise, temporally correlated noise describing benchmark wobble \citep[e.g.,][]{Wyatt1982,Wyatt1989}, and/or spatially correlated noise describing common mode error \citep[e.g.,][]{Wdowinski1997}. For now, we only make the assumption that $\eta_i$ is a Gaussian process distributed as $\eta_i \sim \mathcal{N}(0,C_{\eta_i})$, and $C_{\eta_i}$ is a spatially and temporally dependent covariance function. We consider the coefficients in eq. (\ref{eq:Data}) to be uncorrelated random variables distributed as $\mathcal{N}(0,\kappa^2)$ in the limit as $\kappa \to \infty$ (i.e., the coefficients have diffuse priors). Of course, the secular velocities, $b_i^{(j)}$, are spatially correlated and we could invoke a tectonic model to form a prior on $b_i^{(j)}$. However, in our application to the Pacific Northwest, we will be using displacement time series which are long enough to sufficiently constrain $b_i^{(j)}$ for each station, avoiding the need to incorporate a prior. Likewise, seasonal deformation is spatially correlated \citep{Dong2002,Langbein2008}, and it may be worth exploring and exploiting such a correlation in a future study.

We move on to consider the column vector of $n$ GNSS displacement observations, $\mitbf{d}^*_i$, where the subscript indicates that we are still only considering component $i$ of displacements. The times and positions for each observation are described by the set $\mitbf{P}$. $\mitbf{d}^*_i$ can be considered a realization of the random vector $\mitbf{d}_i$, which is formed by evaluating eq. (\ref{eq:Data}) at each point in $\mitbf{P}$. To write out $\mitbf{d}_i$ explicitly, we collect the coefficients from eq. (\ref{eq:Data}) into the vector $\mitbf{m}_i$, which contains 6 elements for each of the $m$ GNSS station. We use $\mitbf{G}$ to represent the $n \times 6m$ matrix of corresponding basis functions evaluated at each point in $\mitbf{P}$. We can write $\mitbf{d}_i$ as 
\begin{equation}\label{eq:DataVec}
\mitbf{d}_i = u_i(\mitbf{P}) + \eta_i(\mitbf{P}) + \mitbf{G}\mitbf{m}_i,
\end{equation}
which is distributed as $\mathcal{N}(\mitbf{0},\mitbf{\Sigma} + \kappa^2\mitbf{G}\mitbf{G}^T)$, where
\begin{equation}\label{eq:Cd}
\mitbf{\Sigma} = C_{u_i}(\mitbf{P},\mitbf{P}) + C_{\eta_i}(\mitbf{P},\mitbf{P}).  
\end{equation}
It should be understood that notation such as $u_i(\mitbf{P})$ and $C_{u_i}(\mitbf{P},\mitbf{P})$ represents the column vector $[u_i(p)]_{p \in \mitbf{P}}$ and the matrix $[C_{u_i}(p,p')]_{(p,p') \in \mitbf{P} \times \mitbf{P}}$, respectively.

We combine the prior on transient displacements with GNSS displacement data to form a posterior estimate of transient displacements. We ignore any covariances between the easting and northing components, which reduces the computational complexity of our analysis because we can determine each component of the posterior independently. However, we are inherently assuming that the principle components of transient displacements and the noise processes are aligned with the cardinal directions, which is admittedly an arbitrary assumption. We then form each component of the posterior, denoted as $\hat{u}_i$, by conditioning $u_i$ with the fact that $\mitbf{d}^*_i$ has been realized from the random vector $\mitbf{d}_i$. That is to say $\hat{u}_i = u_i |(\mitbf{d}_i = \mitbf{d}^*_i)$. If $\kappa$ is kept finite then, from \citet[sec. 2.2]{Rasmussen2006}, we find that $\hat{u}_i$ is distributed as $\mathcal{N}(\mu_{\hat{u}_i},C_{\hat{u}_i})$, where

@@

\subsection{Outlier detection}\label{sec:Outlier}
In deriving our formulation for transient strain rates, we have assumed that noise in the data vector is normally distributed. This is not an appropriate assumption for GNSS data, which often have more outliers than would be predicted for normally distributed noise. Methods for analyzing GNSS data should either be robust against outliers or should involve a preprocessing step in which outliers are detected and removed. Examples of the former include the MIDAS method for estimating secular velocities \citep{Blewitt2016} and the GPS Imaging method for detecting spatially coherent features \citep{Hammond2016}.  We adopt the latter approach and remove outliers in the GNSS data prior to estimating $\dot\varepsilon$. We identify outliers based on the residuals between the observations and a best fitting model. Observations with residuals that exceed some threshold are removed. This strategy is commonly used to detect outliers in GNSS data, where the residuals are typically computed for each station by estimating and removing a temporally dependent model consisting of a linear trend and seasonal terms \citep[e.g.,][]{Johansson2002,Dong2006,Bos2013}. For the outlier detection method used in this study, the residuals are determined by removing a best fitting spatially and temporally dependent model from the GNSS data. That model that we want to fit to the $n$ GNSS observations, $\mitbf{d}^*$, consists of the linear trend and sinusoids from eq. (\ref{eq:DataVec}), $\mitbf{G}\mitbf{m}$, and a Gaussian process, $v(\mitbf{P})$, that is used to describe any spatially and temporally correlated features in the GNSS data. We now consider $\mitbf{d}^*$ to be a realization of the random vector
\begin{equation}
\mitbf{d} = \mitbf{G}\mitbf{m} + v(\mitbf{P}) + \mitbf{w},
\end{equation}   
where $\mitbf{w}$ is a vector of uncorrelated Gaussian noise distributed as $\mitbf{w} = \mathcal{N}\left(0,\mathrm{diag}(\mitbf{\sigma}^2)\right)$, and $\mitbf{\sigma} = [\sigma_1, \sigma_2, ..., \sigma_n]$ are the data uncertainties. The residual vector, $\mitbf{r} = [r_1,r_2,...,r_n]$, is the difference between $\mitbf{d}^*$, and the mean of the $\mitbf{G}\mitbf{m} + v(\mitbf{P})$ conditioned on $\mitbf{d} = \mitbf{d}^*$.

This best fitting model is found using Gaussian process regression, where the model consists of a parametric and

We again use Gaussian process regression to find the best fitting model, which consists of parametric and non-parametric

We now describe column vector of $n$ GNSS observations $\mitbf{d}*$ to be a

With this approach, outliers are considered to by any observation that is spatially or temporally anomalous.

approach for identifying outliers differs

is involving fitting a model to , in principle, similar to

Methods for analyzing GNSS data should either be robust against outliers or a outliers , such as the MIDAS method for estimating secular velocities \citep{Blewitt2016}, or

It follows that proposed methods for analyzing GNSS data should be robust against outliers \citep[e.g.,][]{Blewitt2016}. In order to make our estimates of transient strain more robust, we automatically identify and remove outliers in the GNSS data as a pre-processing step.
